Title Goes Here\\ \large{Not For Distribution}

### 2020-03-22 paper/paper.tex
Some updates to the paper

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Title Goes Here\\
\large{Not For Redistribution}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Domain specific languages (DSLs) provide a powerful tool for to abstract over
  a class of problems. In this paper, we introduce a technique for representing
  algebraic data types and pattern matching in a DSL embedded in Haskell. This
  representation is mechanized through the use of GHC Generics and and GHC Core plugin.
\end{abstract}

@@

\begin{itemize}
\item A representation of pattern matching in a Haskell EDSL (Section 3)
\item Representations of tail recursion (Section 4) and lambdas (Section 5) in
  the EDSL. We will see that lambdas become easier to implement once pattern
  matching and tail recursion already have EDSL representations.
\item A GHC Core plugin which transforms a subset of standard Haskell code
  into this EDSL for pattern matching, extended with tail recursion,
  lambdas and primitive operations (Section 8)
\item An implementation of a interpreter for the "standard" semantics of the EDSL (Section 6).
\end{itemize}

@@

@@

example' :: Int
example' =
  abs (CaseExp
        (rep x)
        (SumMatchExp (OneProdMatch
                       (Lam 1
                         (Ord (Var 1))))
                     (OneSumMatch
                       (OneProdMatch
                         (Lam 2
                           (Var 2)))))
\end{lstlisting}

@@

@@

If \ttt{ERepTy ComplexPair} were fully deconstructed into \ttt{((Double, Double),
(Double, Double))}, we would need a \ttt{Num} instance for \ttt{(Double,
Double)}.  What we really want is to use the fact that a \ttt{Num} instance
already exists for \ttt{Complex Double}. This is exactly what preserving this
type information allows us to do.

@@

This information is brought into the \ttt{E} type via the constructor
\ttt{Repped}. The \ttt{E} also contains constructors representing \ttt{Either}
and \ttt{(,)} values:

\begin{lstlisting}
data E t where
  ...
  Repped :: ERep t => ERepTy t -> E t

@@

\begin{lstlisting}
class ERep t where
  type ERepTy t

  rep :: t -> E t
  rep = Repped . rep'

  unrep :: ERepTy t -> t

@@

  unrep = id
  rep'  = id

@@

  rep (x, y) = PairExp (rep x) (rep y)
  unrep = id
  rep'  = id
\end{lstlisting}

@@

The overall structure of a \ttt{E (SumMatch a b)} value is a (heterogeneous)
list of \ttt{E (ProdMatch x y)} values.  Each item of this list corresponds
exactly to one branch in the original \ttt{case} match.

@@

absEnv env (Repped x) = unrep x

@@

In the following flowchart, the names in parentheses in a node refers to the names of transformations from the following section (Section 8.2).

\begin{tikzpicture}[node distance=2cm]
\node (start) [startstop] {Core from GHC};
  \node (tailrec) [process, below of=start, yshift=-0.5cm]{($D, T_{\ttt{f}}, T'_{\ttt{f}}$) Transform tail recursion of definitions that have a mark (when applicable)};
  \node (findmark) [process, below of=tailrec, yshift=-0.5cm]{($C^*$) Find marked expression and erase mark};
    \node (handlecase) [process, below of=findmark]{($C, M$) Mark scrutinee};
    \node (handleprim) [process, right of=handlecase, xshift=1.6cm]{($C$) Encode as EDSL primitive};
    \node (handleapp) [process, right of=handleprim, xshift=1.6cm]{($C$) Wrap with \ttt{Construct} and \ttt{ConstructAp}};

    \node (marksubs1) [process, below of=handleprim]{($C, M$) Mark sub-expressions};
    \node (marksubs2) [process, below of=handleapp]{($C, M$) Mark sub-expressions};

@@

  \draw [arrow] (start) -- (tailrec);
  \draw [arrow] (tailrec) -- (findmark);
  \draw [arrow] (findmark) -- node[anchor=east] {Primitive} (handleprim);
  \draw [arrow] (findmark) -- node[anchor=west, yshift=0.1cm] {Function application/non-prim var} (handleapp);
  \draw [arrow] (findmark) -- node[anchor=east] {\ttt{case}} (handlecase);

@@

  \draw [arrow] (handleapp) -- (marksubs2);

@@

  \draw [arrow] (marksubs1) -- (endstep);
  \draw [arrow] (marksubs2) -- (endstep);
  \draw [arrow] (markbody) -- (endstep);

@@

\clearpage
\subsection{Syntactic Transformations of the Core Plugin}
\begin{todo}
  Should we use Oxford brackets here? Maybe a bit unusual, but we are operating syntactically and mapping from \ttt{a} to \ttt{E a} (could \ttt{E a} be considered an "intermediary" semantic domain?).
\end{todo}
\vspace*{-0.6cm}
\begin{align*}
  D&\expr{\ttt{f x = internalize (externalize (case s of \{ ... \}))}}\\
      &\rewrites \begin{cases}
        \ttt{f = internalize $C^{*}\expr{\ttt{externalize (runIter ($\lambda \ttt{x} \rarr \ttt{case s of \{ $T'_{\ttt{f}}\exprdots$}$ \}))}}$} & \text{if \ttt{f} occurs free in \ttt{\{ ... \}}}\\
        \ttt{f x = internalize $C^{*}\expr{\ttt{externalize (case s of \{ ... \})}}$} & \text{otherwise}
      \end{cases}
\end{align*}

\begin{align*}
  T'_{\ttt{f}}&\expr{\ttt{K x0 ...$_U$ xN $\rarr$ case s of \{ ...$_V$ \}; ...$_W$}}\\
    &\rewrites
        \ttt{K x0 ...$_U$ xN $\rarr$ $T_{\ttt{f}}\expr{\ttt{case s of \{ ...$_V$ \}}}$; $T'_{\ttt{f}}\expr{\ttt{ ...$_W$ }}$}\\
  T'_{\ttt{f}}&\expr{\ttt{K x0 ...$_U$ xN $\rarr$ body0; ...$_V$ }}\\
    &\rewrites
      \begin{cases}
        \ttt{K x0 ...$_U$ xN $\rarr$ body0}[\ttt{f} \mapsto \ttt{Step}]\ttt{; }T'_{\ttt{f}}\expr{\ttt{ ...$_V$ }} & \text{if \ttt{f} occurs free in \ttt{body0}}\\
        \ttt{K x0 ...$_U$ xN $\rarr$ Done body0; $T'_{\ttt{f}}\expr{\ttt{ ...$_V$ }}$} & \text{otherwise}
      \end{cases}\\
  T'_{\ttt{f}}&\expr{\varepsilon} \rewrites \varepsilon\;\;\; \text{(Note: $\varepsilon$ represents the empty string)}\\
  T_{\ttt{f}}&\expr{\ttt{case s of \{ ... \}}}\\
    &\rewrites
      \begin{cases}
        \ttt{case s of \{ $T'_{\ttt{f}}\expr{\ttt{ ... }}$ \}} & \text{if \ttt{f} occurs free in \ttt{\{ ... \}}}\\
        \ttt{Done (case s of \{ ... \})} & \text{otherwise}
      \end{cases}
\end{align*}
\begin{align*}
  &C^{*}\expr{\ttt{x}} \rewrites
    \begin{cases}
      \ttt{x} & \text{if \ttt{x} has no subexpressions marked with \ttt{externalize}}\\
      C^{*}\expr{C\expr{\ttt{y}}} & \text{if \ttt{externalize y} is the first marked subexpression of \ttt{x}}\\
    \end{cases}\\
\end{align*}
\begin{align*}
  &C\expr{\ttt{runIter x}} \rewrites \ttt{TailRec $M\expr{\ttt{x}}$}\\
  &C\expr{\ttt{x + y}} \rewrites \ttt{Add $M\expr{\ttt{x}}$ $M\expr{\ttt{y}}$}\\
  &C\expr{\ttt{case scrutinee of \{ ... \}}} \rewrites \ttt{CaseExp $M\expr{\ttt{scrutinee}}$ } B\expr{\ttt{ ... }}\\
  &C\expr{\ttt{$\lambda$x $\rarr$ body}} \rewrites A\expr{\lambda \ttt{x} \rarr \ttt{body}}\\
  &C\expr{\ttt{f x}} \rewrites \ttt{ConstructAp $M\expr{\ttt{f}}$ $M\expr{\ttt{x}}$}\;\;\;\;\text{(Where \ttt{f} is not a primitive or an application of a primitive)}\\
  &C\expr{\ttt{x}} \rewrites \ttt{Construct x}\;\;\;\;\text{(Where \ttt{x} is a variable which is not a primitive)}\\
  &\\
  &B\expr{\ttt{K x0 ...$_U$ xN $\rarr$ body0; ...$_V$}} \rewrites \ttt{SumMatchExp }P\expr{\ttt{x0 ...$_U$ xN $\rarr$ body0}}\; B\expr{\ttt{ ...$_V$ }}\\
  &B\expr{\ttt{K x0 ... xN $\rarr$ body}} \rewrites \ttt{OneSumMatchExp }P\expr{\ttt{x0 ... xN $\rarr$ body}}\\
  &B\expr{\ttt{$\varepsilon$}} \rewrites \ttt{EmptyMatch}\\
  &\\
  &P\expr{\ttt{x0 x1 ... $\rarr$ body}} \rewrites \ttt{ProdMatchExp }A\expr{\ttt{$\lambda$ x0 $\rarr$ $P\expr{\ttt{x1 ... $\rarr$ body}}$}}\\
  &P\expr{\ttt{x $\rarr$ body}} \rewrites \ttt{OneProdMatchExp }A\expr{\ttt{$\lambda$ x $\rarr$ body}}\\
  &P\expr{\ttt{ $\rarr$ body}} \rewrites \ttt{NullaryMatch $M\expr{\ttt{body}}$}\\
  &\\
  &A\expr{\ttt{$\lambda$(x :: a) $\rarr$ body}} \rewrites \ttt{Lam (Name @a uniq) $M\expr{\ttt{body$[\ttt{x} \mapsto \ttt{Var uniq}]$}}$}\text{\;\;\;\;\;(where \ttt{uniq} is a globally unique identifier)}\\
  &\\
  &M\expr{\ttt{x :: a}} \rewrites
    \begin{cases}
      \ttt{externalize x} & \text{if}\; \nexists t, a\; \typeeq E\; t\\
      \ttt{x} & \text{if}\; \exists t, a\; \typeeq E\; t
    \end{cases}
\end{align*}

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Title Goes Here\\
\large{Not For Distribution}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Domain specific languages (DSLs) provide a powerful tool for abstracting over
  a class of problems. In this paper, we introduce a technique for representing
  algebraic data types and pattern matching in a DSL embedded in Haskell. This
  representation is mechanized through the use of GHC Generics and and GHC Core plugin.
\end{abstract}

@@

\begin{itemize}
\item A representation of pattern matching in a Haskell EDSL (Section 3)
\item A representation of lambdas (Section 5) in the EDSL, taking advantage of
  the fact that both pattern matching and tail recursion are already captured in
  the EDSL.
\item A GHC Core plugin which transforms a subset of standard Haskell code
  into this EDSL for pattern matching, extended with tail recursion,
  lambdas and primitive operations (Section 8)
\item An implementation of a interpreter for the "standard" semantics of the EDSL (Section 6).
\item An outline of a C backend for this EDSL (Section 7).
\end{itemize}

@@

Additionally, the following functions mark the parts of the code which the Core
plugin will target:

\begin{lstlisting}
internalize :: E a -> a
externalize :: a -> E a
\end{lstlisting}

@@

example' :: Int
example' =
  abs (CaseExp
        (LeftExp 'a')
        (SumMatchExp (OneProdMatch
                       (Lam 1
                         (Ord (Var 1))))
                     (OneSumMatch
                       (OneProdMatch
                         (Lam 2
                           (Var 2)))))
\end{lstlisting}

@@

\begin{todo}
  Look into \ttt{generic-sop} and compare then cite corresponding paper.
\end{todo}

@@

If \ttt{ERepTy ComplexPair} were fully deconstructed into \ttt{((Double, Double),
(Double, Double))}, we would need a \ttt{Num} instance for \ttt{(Double,
Double)}.  What we really want is to use the fact that a \ttt{Num} instance
already exists for \ttt{Complex Double}. This is exactly what preserving this
type information allows us to do. We can later use this preserved type
information in backends, through the corresponding \ttt{Typeable} instances (for
instance, to provide special support for arithmetic on complex numbers).

@@

This information is brought into the \ttt{E} type via the constructor
\ttt{ConstructRep}. The \ttt{E} also contains constructors representing \ttt{Either}
and \ttt{(,)} values:

\begin{lstlisting}
data E t where
  ...
  ConstructRep ::
    (Typeable a, ERep a)
    => E (ERepTy a) -> E a

@@

\begin{lstlisting}
class Typeable t => ERep t where
  type ERepTy t

  construct :: t -> E (ERepTy t)

  rep :: t -> E t

  default rep ::
    (ERep (ERepTy t))
    => t -> E t
  rep x = ConstructRep (construct x)
  {-# INLINABLE rep #-}

  unrep' :: ERepTy t -> t

@@

  unrep' = id
  rep'   = id

@@

  rep (x, y) = PairExp (rep x) (rep y)
  unrep' = id
  rep'   = id
\end{lstlisting}

@@

The overall structure of a \ttt{E (SumMatch a b)} value is a (heterogeneous)
list of \ttt{E (ProdMatch x b)} values.  Each item of this list corresponds
exactly to one branch in the original \ttt{case} match.

@@

absEnv env (ConstructRep x) =
  unrep' (absEnv env x)

@@

In the following flowchart, the names in parentheses in a node refers to the names of transformations from the following section (Section 8.2). Note that
calls to \ttt{unrep :: E a -> a} are only used internally and will not exist in the final result of the transformation.

\begin{tikzpicture}[node distance=2cm]
\node (start) [startstop] {Core from GHC};
  \node (tailrec) [process, below of=start, yshift=-0.5cm]{($D, T_{\ttt{f}}, T'_{\ttt{f}}$) Transform tail recursion of definitions that have a mark (when applicable)};
  \node (findmark) [process, below of=tailrec, yshift=-0.5cm]{($C^*$) Find marked expression and erase mark};
    \node (handlecase) [process, below of=findmark]{($C, M$) Mark scrutinee};
    \node (handleprim) [process, right of=handlecase, xshift=1.6cm]{($C$) Encode as EDSL primitive};
    \node (handleconstr) [process, right of=handleprim, xshift=1.6cm]{($C$) Wrap with \ttt{construct}};

    \node (marksubs1) [process, below of=handleprim]{($C, M$) Mark sub-expressions};
    \node (marksubs2) [process, below of=handleconstr]{($C, M$) Mark sub-expressions and wrap with \ttt{unrep}};
    \node (betareduce) [process, below of=marksubs2]{Beta reduce \ttt{(construct ...)} expression and simplify};
    \node (repunrep) [process, below of=betareduce]{Replace expressions of the form \ttt{(rep (unrep x))} with \ttt{x}};

@@

  \draw [arrow] (start) -- (tailrec);
  \draw [arrow] (tailrec) -- (findmark);
  \draw [arrow] (findmark) -- node[anchor=east] {Primitive} (handleprim);
  \draw [arrow] (findmark) -- node[anchor=west, yshift=0.1cm] {Constructor/constructor application} (handleconstr);
  \draw [arrow] (findmark) -- node[anchor=east] {\ttt{case}} (handlecase);

@@

  \draw [arrow] (handleconstr) -- (marksubs2);

@@

  \draw [arrow] (marksubs2) -- (betareduce);
  \draw [arrow] (betareduce) -- (repunrep);

  \draw [arrow] (marksubs1) -- (endstep);
  \draw [arrow] (repunrep) -- (endstep);
  \draw [arrow] (markbody) -- (endstep);

@@

\clearpage
\subsection{Syntactic Transformations of the Core Plugin}

% \begin{todo}
%   Should we use Oxford brackets here? Maybe a bit unusual, but we are operating syntactically and mapping from \ttt{a} to \ttt{E a} (could \ttt{E a} be considered an "intermediary" semantic domain?).
% \end{todo}
% \vspace*{-0.6cm}

\begin{align*}
  D&\expr{\ttt{f x = internalize (externalize (case s of \{ ... \}))}}\\
      &\rewrites \begin{cases}
        \ttt{f = internalize $C^{*}\expr{\ttt{externalize (runIter ($\lambda \ttt{x} \rarr \ttt{case s of \{ $T'_{\ttt{f}}\exprdots$}$ \}))}}$} & \text{if \ttt{f} occurs free in \ttt{\{ ... \}}}\\
        \ttt{f x = internalize $C^{*}\expr{\ttt{externalize (case s of \{ ... \})}}$} & \text{otherwise}
      \end{cases}
\end{align*}

\begin{align*}
  T'_{\ttt{f}}&\expr{\ttt{K x0 ...$_U$ xN $\rarr$ case s of \{ ...$_V$ \}; ...$_W$}}\\
    &\rewrites
        \ttt{K x0 ...$_U$ xN $\rarr$ $T_{\ttt{f}}\expr{\ttt{case s of \{ ...$_V$ \}}}$; $T'_{\ttt{f}}\expr{\ttt{ ...$_W$ }}$}\\
  T'_{\ttt{f}}&\expr{\ttt{K x0 ...$_U$ xN $\rarr$ body0; ...$_V$ }}\\
    &\rewrites
      \begin{cases}
        \ttt{K x0 ...$_U$ xN $\rarr$ body0}[\ttt{f} \mapsto \ttt{Step}]\ttt{; }T'_{\ttt{f}}\expr{\ttt{ ...$_V$ }} & \text{if \ttt{f} occurs free in \ttt{body0}}\\
        \ttt{K x0 ...$_U$ xN $\rarr$ Done body0; $T'_{\ttt{f}}\expr{\ttt{ ...$_V$ }}$} & \text{otherwise}
      \end{cases}\\
  T'_{\ttt{f}}&\expr{\varepsilon} \rewrites \varepsilon\;\;\; \text{(Note: $\varepsilon$ represents the empty string)}\\
  T_{\ttt{f}}&\expr{\ttt{case s of \{ ... \}}}\\
    &\rewrites
      \begin{cases}
        \ttt{case s of \{ $T'_{\ttt{f}}\expr{\ttt{ ... }}$ \}} & \text{if \ttt{f} occurs free in \ttt{\{ ... \}}}\\
        \ttt{Done (case s of \{ ... \})} & \text{otherwise}
      \end{cases}
\end{align*}
\begin{align*}
  &C^{*}\expr{\ttt{x}} \rewrites
    \begin{cases}
      \ttt{x} & \text{if \ttt{x} has no subexpressions marked with \ttt{externalize}}\\
      C^{*}\expr{C\expr{\ttt{y}}} & \text{if \ttt{externalize y} is the first marked subexpression of \ttt{x}}\\
    \end{cases}\\
\end{align*}
\begin{align*}
  &C\expr{\ttt{runIter x}} \rewrites \ttt{TailRec $M\expr{\ttt{x}}$}\\
  &C\expr{\ttt{x + y}} \rewrites \ttt{Add $M\expr{\ttt{x}}$ $M\expr{\ttt{y}}$}\\
  &C\expr{\ttt{case scrutinee of \{ ... \}}} \rewrites \ttt{CaseExp $M\expr{\ttt{scrutinee}}$ } B\expr{\ttt{ ... }}\\
  &C\expr{\ttt{$\lambda$x $\rarr$ body}} \rewrites A\expr{\lambda \ttt{x} \rarr \ttt{body}}\\
  &C\expr{\ttt{K x0 ... xN }} \rewrites \ttt{construct (K (unrep $M\expr{\ttt{x0}}$) ... (unrep $M\expr{\ttt{xN}}$))}\;\;\;\;\text{(Where \ttt{K} is a constructor)}\\
  % &C\expr{\ttt{f x}} \rewrites \ttt{ConstructAp $M\expr{\ttt{f}}$ $M\expr{\ttt{x}}$}\;\;\;\;\text{(Where \ttt{f} is not a primitive or an application of a primitive)}\\
  % &C\expr{\ttt{x}} \rewrites \ttt{Construct x}\;\;\;\;\text{(Where \ttt{x} is a variable which is not a primitive)}\\
  &\\
  &B\expr{\ttt{K x0 ...$_U$ xN $\rarr$ body0; ...$_V$}} \rewrites \ttt{SumMatchExp }P\expr{\ttt{x0 ...$_U$ xN $\rarr$ body0}}\; B\expr{\ttt{ ...$_V$ }}\\
  &B\expr{\ttt{K x0 ... xN $\rarr$ body}} \rewrites \ttt{OneSumMatchExp }P\expr{\ttt{x0 ... xN $\rarr$ body}}\\
  &B\expr{\ttt{$\varepsilon$}} \rewrites \ttt{EmptyMatch}\\
  &\\
  &P\expr{\ttt{x0 x1 ... $\rarr$ body}} \rewrites \ttt{ProdMatchExp }A\expr{\ttt{$\lambda$ x0 $\rarr$ $P\expr{\ttt{x1 ... $\rarr$ body}}$}}\\
  &P\expr{\ttt{x $\rarr$ body}} \rewrites \ttt{OneProdMatchExp }A\expr{\ttt{$\lambda$ x $\rarr$ body}}\\
  &P\expr{\ttt{ $\rarr$ body}} \rewrites \ttt{NullaryMatch $M\expr{\ttt{body}}$}\\
  &\\
  &A\expr{\ttt{$\lambda$(x :: a) $\rarr$ body}} \rewrites \ttt{Lam (Name @a uniq) $M\expr{\ttt{body$[\ttt{x} \mapsto \ttt{Var uniq}]$}}$}\text{\;\;\;\;\;(where \ttt{uniq} is a globally unique identifier)}\\
  &\\
  &M\expr{\ttt{x :: a}} \rewrites
    \begin{cases}
      \ttt{externalize x} & \text{if}\; \nexists t, a\; \typeeq E\; t\\
      \ttt{x} & \text{if}\; \exists t, a\; \typeeq E\; t
    \end{cases}
\end{align*}
